Data Specialization for a Staged Lambda Calculus

### 2015-05-16 paper/semantics.tex
more work on small-step semantics

\subsection{(Small Step) Dynamics}
We provide a small-step semantics which evaluates \lang\ terms in two passes,
first reducing only redexes in the first stage portions of code,
then reducing those in second stage portions of code.
Because first- and second-stage code are
interleaved in \lang\ terms, realizing this notion of staged
evaluation requires some care.

[Example without hoisting.]
[Paragraph about why we want hoisting.]
[Gist of hoisting solution.]

The first pass of evaluation comprises three judgements: 
steps $\stepwor e e'$, which asserts that a term $e$ performs some internal reduction to produce $e'$,
where both terms are at world $w$;
doneness $e \done$, which asserts that a term $e$ at world $w$ has no more first stage redexes;
and hoisting steps $\lift e {e'}$, which are available only at world \bbonem.

@@

\paragraph{Done.} The judgement $e \done$ indicates that a term $e$ contains no first stage redexes.
In this way, it generalizes over the notions of values and residuals.

>>>>

\subsection{(Small Step) Dynamics}
We provide a small-step semantics which,
given a \lang\ term $e$, evaluates its first stage subterms
either until $e$ is a value (if $e$ is at world \bbonep\ or \bbonem)
of until all of $e$'s first stage subterms have been evaluated away
(when $e$ is at world \bbtwo).
Because first- and second-stage code are
interleaved in \lang\ terms, this first pass of
evaluation requires some care.

Moreover, our staged semantics contains one innovation over those of 
\cite{davies96} and \cite{taha-thesis-99}, which is the addition
of hoisting steps.  
To illustrate, consider the example
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+4}}
\end{lstlisting}
A non-staged semantics which ignores $\next$s and $\prev$s
would proceed on this term by reducing \verb|2+3| to \verb|6| in place.
A stage semantics does not, because it only reduces first stage redexes, and
\verb|2+3| is at the second stage.  
The staged semantics of \cite{davies96} and \cite{taha-thesis-99} instead proceed by
substituting \verb|next{2+3}| in for \verb|x| to yield
\begin{lstlisting}
prev{next{prev{next{2*3}}+4}}
\end{lstlisting}
This takes one step to 
\begin{lstlisting}
prev{next{2*3+4}}
\end{lstlisting}
and then one more step to
\begin{lstlisting}
2*3+4
\end{lstlisting}
which is the final entirely-second-stage result, called a {\em residual}.
Our semantics differs from this formulation in one way.
Consider a slightly modified example where \verb|x| appears twice:
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+prev{x}}}
\end{lstlisting}
A standard stage semantics would, in four steps, evaluate this to 
\begin{lstlisting}
2*3+2*3
\end{lstlisting}
Note how the multiplication is duplicated in the residual.
We find this behavior potentially undesirable for our applications,
so our semantics instead {\em hoists} the contents of a next statement 
to the outermost containing $\prev$, turning
\begin{lstlisting}
prev{let x = next{2*3} in next{prev{x}+prev{x}}}
\end{lstlisting}
into
\begin{lstlisting}
let y = 2*3 in
prev{let x = next{y} in next{prev{x}+prev{x}}}
\end{lstlisting}
This then evaluates to 
\begin{lstlisting}
let y = 2*3 in y+y
\end{lstlisting}
in four steps.  
This hoisting step occurs as the last step in the evaluation of a $\next$.

Formally, the dynamic semantics comprises three judgements: 
steps $\stepwor e e'$, which asserts that a term $e$ performs some internal reduction to produce $e'$,
where both terms are at world $w$;
doneness $e \done$, which asserts that a term $e$ at world $w$ has no more first stage redexes;
and hoisting steps $\lift e {e'}$, which are available only at world \bbonem.

@@

\paragraph{Done.} The judgement $e \done$ indicates that a term $e$ contains no first stage redexes.
It generalizes over the notions of values and residuals.
